---
title: "R Notebook"
output: html_notebook
---


```{r}
if("DBI" %in% rownames(installed.packages()) == FALSE) {
  install.packages("DBI")  
}
library("DBI")

if("RSQLite" %in% rownames(installed.packages()) == FALSE) {
  install.packages("RSQLite")  
}
library("sqldf")

if("sqldf" %in% rownames(installed.packages()) == FALSE) {
  install.packages("sqldf")  
}
library("sqldf")

if("readxl" %in% rownames(installed.packages()) == FALSE) {
  install.packages("readxl")  
}
library("readxl")
library("stringr")
library(tidyverse)
options(sqldf.driver = 'SQLite')
```



Loading raw health & econ databases
```{r}
health <- na.omit(read.csv("HealthData.csv"))
WB_DBI <- na.omit(read.csv("WB_WDI.csv"))

worldBank <- WB_DBI %>% select(
                  location = Country.Name, 
                  series = Series.Name,
                  `2000` = X2000..YR2000., 
                  `2005` = X2005..YR2005.,
                  `2010` = X2010..YR2010.,
                  `2015` = X2015..YR2015.,
                  `2019` = X2019..YR2019.) %>%
  pivot_longer(`2000`:`2019`, names_to = "year") 


worldBank$value[worldBank$value == ".."] <- NA

worldBank$location <- str_squish(gsub("\\(.*", "", worldBank$location))
health$location <- str_squish(gsub("\\(.*", "", health$location))
```

These names are in econ data but not in health data
```{r}
countriesW <- unique(worldBank$location)
countriesH <- unique(health$location)
wcountries <- countriesW[!countriesW %in% countriesH]
wcountries[order(wcountries)]
```

These names are in health but not in econ data
```{r}
countriesW <- unique(worldBank$location)
countriesH <- unique(health$location)
hcountries <- countriesH[!countriesH %in% countriesW]
hcountries[order(hcountries)]

```

Normalizing the names of any countries in both databases:

```{r}
worldBank$location[worldBank$location == "Bahamas, The"] <- "Bahamas"
worldBank$location[worldBank$location == "Korea, Dem. People's Rep."] <- "Democratic People's Republic of Korea"
worldBank$location[worldBank$location == "Korea, Rep."] <- "Republic of Korea"
worldBank$location[worldBank$location == "Congo, Dem. Rep."] <- "Democratic Republic of the Congo"
worldBank$location[worldBank$location == "Egypt, Arab Rep."] <- "Egypt" 
worldBank$location[worldBank$location == "Gambia, The"] <- "Gambia"
worldBank$location[worldBank$location == "Iran, Islamic Rep."] <- "Iran"
worldBank$location[worldBank$location == "Kyrgyz Republic"] <- "Kyrgyzstan"
worldBank$location[worldBank$location == "Congo, Rep."] <- "Congo"
worldBank$location[worldBank$location == "Micronesia, Fed. Sts."] <- "Micronesia"
worldBank$location[worldBank$location == "West Bank and Gaza"] <- "Palestine"
worldBank$location[worldBank$location == "Slovak Republic"] <- "Slovakia"
worldBank$location[worldBank$location == "Turkiye"] <- "Turkey"
worldBank$location[worldBank$location == "Virgin Islands"] <- "United States Virgin Islands"
worldBank$location[worldBank$location == "Venezuela, RB"] <- "Venezuela"
worldBank$location[worldBank$location == "Yemen, Rep."] <- "Yemen"

health$location[health$location == "Republic of Moldova"] <- "Moldova"
health$location[health$location == "United States of America"] <- "United States"
health$location[health$location == "Côte d'Ivoire"] <- "Cote d'Ivoire"
health$location[health$location =="Lao People's Democratic Republic"] <- "Lao PDR"
health$location[health$location =="Saint Kitts and Nevis"] <- "St. Kitts and Nevis"
health$location[health$location =="Saint Lucia"] <- "St. Lucia"
health$location[health$location =="Saint Vincent and the Grenadines"] <- "St. Vincent and the Grenadines"
health$location[health$location =="United Republic of Tanzania"] <- "Tanzania"
health$location[health$location =="Viet Nam"] <- "Vietnam"

```

Eliminating all countries without a match now

```{r}
countriesW <- unique(worldBank$location)
countriesH <- unique(health$location)
wcountries <- countriesW[!countriesW %in% countriesH]
hcountries <- countriesH[!countriesH %in% countriesW]

worldBank <- worldBank[!(worldBank$location %in% wcountries),]
health <- health[!(health$location %in% hcountries),]
print("Deleted Countries:")
print(hcountries)
print(wcountries)
```


Count how many nulls each Series has and the percent missing of total series
```{r}
series <- unique(worldBank$series[worldBank$series != ""])

null.count <- data.frame(Series = character(length(series)))
for(j in 1:length(series)){
  null.count$NullCount[j] <- 0
  null.count$PercentMissing[j] <- 0
}

j = 1
for(i in 1:length(series)){
  col <-worldBank[worldBank$series == series[i],]
  null.count$Series[i] = series[i]
  null.count$NullCount[i] = sum(is.na(col$value))
  null.count$PercentMissing[i] = sum(is.na(col$value)) / nrow(col) * 100

}

null.count <- null.count[order(null.count$NullCount),]
null.count
```

Eliminate all series with more than 10% missing:
```{r}
seriesList <- null.count$Series[null.count$PercentMissing > 10]
seriesList

economic.10p <- worldBank[!worldBank$series %in% seriesList,]
economic.10p
```

Count how many nulls each country has and the percent missing
```{r}
locations <- unique(economic.10p$location[economic.10p$location != ""])

null.count1 <- data.frame(Countries = character(length(locations)))
for(j in 1:length(locations)){
  null.count1$NullCount[j] <- 0
  null.count1$PercentMissing[j] <- 0
}


for(i in 1:length(locations)){
  col <-economic.10p[economic.10p$location == locations[i],]
  null.count1$Countries[i] = locations[i]
  null.count1$NullCount[i] = sum(is.na(col$value))
  null.count1$PercentMissing[i] = sum(is.na(col$value)) / nrow(col) * 100
}

null.count1 <- null.count1[order(null.count1$NullCount),]
null.count1
```

Total missing:
With these 200 countries, 4.25% of our data values are null, but there are a few countries that are missing A LOT of data points.
If countries are each an observation, I imagine we could eliminate some as long as it doesn't skew the data in some specific direction
```{r}
sum(is.na(economic.10p$value)) / length(economic.10p$value) * 100
```

Eliminate all countries with more than 10% missing:
```{r}
locList <- null.count1$Countries[null.count1$PercentMissing > 10]
locList

economic.10p <- economic.10p[!economic.10p$location %in% locList,]
health <- health[!health$location %in% locList,]
```
There are 22 countries in the data set that had more then 10% null values. Several of these are territories and not necessarily countries, so I think we could eliminate some (if not all) of them:

These are the stats when the 22 countries are taken out

Total percentage of nulls goes down to less than 1%
Total missing: .76%
```{r}
sum(is.na(economic.10p$value)) / length(economic.10p$value) * 100
```

Total countries: 178
```{r}
length(unique(economic.10p$location))
length(unique(health$location))
```

Total Series: 39
```{r}
length(unique(economic.10p$series))
```

Total Datapoints: 34710
```{r}
length(economic.10p$value)
```


# Impute NA values:

```{r}
mean_values <- ave(economic$Data, economic$Location, economic$Series, FUN = function(x) mean(x, na.rm=TRUE))

economic_i <- economic
economic_i$Data <- ifelse(is.na(economic_i$Data), mean_values, economic_i$Data)
```

# Database build starts here

Database will be built from these two data frames:
```{r}
economic.10p$value <- as.numeric(economic.10p$value)

mean_values <- ave(economic.10p$value, economic.10p$location, economic.10p$series, FUN = function(x) mean(x, na.rm=TRUE))
economic.10p$value <- ifelse(is.na(economic.10p$value), mean_values, economic.10p$value)
```

Build data frames for each table
```{r}
db.conn <- dbConnect(RSQLite::SQLite(), dbname="worldHealth_v1.db")

df.yearlyGBD <- sqldf::sqldf(
  "select 1 as GBid,
  cause as Cause,
  val as Data,
  location as Location,
  year as Year
  from `health`"
)

df.yearlyWB <- sqldf::sqldf("
  select 1 as WBid,
  series as Series,
  value as Data,
  location as Location,
  year as Year
  from `economic.10p`")

df.locations <- sqldf::sqldf("
  select 1 as lid, 
  Location 
  from `df.yearlyGBD` 
  group by Location
  ")

df.years <- sqldf::sqldf(
  "select 1 as yid,
  year as Year
  from `df.yearlyGBD`
  group by Year
  ")

df.causes <- sqldf::sqldf(
  "select 1 as cid,
  cause as Cause,
  metric as Metric
  from `health`
  group by Cause"
)

df.series <- sqldf::sqldf(
  "select 1 as sid,
  series as Series,
  series as Metric
  from `economic.10p`
  group by Series"
)
```


Set foreign keys
```{r}
n.locations <- nrow(df.locations)
df.locations[,1] <- seq(1, n.locations)

n.years <- nrow(df.years)
df.years[,1] <- seq(1, n.years)

n.causes <- nrow(df.causes)
df.causes[,1] <- seq(1, n.causes)

n.series <- nrow(df.series)
df.series[,1] <- seq(1, n.series)

n.yearlyGBD <- nrow(df.yearlyGBD)
df.yearlyGBD[,1] <- seq(1, n.yearlyGBD)

n.yearlyWB <- nrow(df.yearlyWB)
df.yearlyWB[,1] <- seq(1, n.yearlyWB)

for(r in 1:n.yearlyGBD){
  # set location id
  lid <- df.locations$lid[which(df.locations$Location == df.yearlyGBD$Location[r])]
  df.yearlyGBD$Location[r] <- lid

  # set series id
  cid <- df.causes$cid[which(df.causes$Cause == df.yearlyGBD$Cause[r])]
  df.yearlyGBD$Cause[r] <- cid
  
  # set year id
  yid <- df.years$yid[which(df.years$Year == df.yearlyGBD$Year[r])]
  df.yearlyGBD$Year[r] <- yid

}

for(j in 1:n.yearlyWB){
  # set location id
  lid <- df.locations$lid[which(df.locations$Location == df.yearlyWB$Location[j])]
  df.yearlyWB$Location[j] <- lid
    
  # set series id
  sid <- df.series$sid[which(df.series$Series == df.yearlyWB$Series[j])]
  df.yearlyWB$Series[j] <- sid
  
  # set year id
  yid <- df.years$yid[which(df.years$Year == df.yearlyWB$Year[j])]
  df.yearlyWB$Year[j] <- yid

}

```

Write tables to db
```{r}
dbExecute(db.conn,"
          DROP TABLE IF EXISTS yearlyGDB")
dbExecute(db.conn,"
          DROP TABLE IF EXISTS yearlyWB")
dbExecute(db.conn,"
          DROP TABLE IF EXISTS years")
dbExecute(db.conn,"
          DROP TABLE IF EXISTS causes")
dbExecute(db.conn,"
          DROP TABLE IF EXISTS locations")
dbExecute(db.conn,"
          DROP TABLE IF EXISTS series")

dbExecute(db.conn,"
          CREATE TABLE years  
          (
            Yid INTEGER PRIMARY KEY,
            Name TEXT
          )")

dbExecute(db.conn,"
          CREATE TABLE locations  
          (
            Lid INTEGER PRIMARY KEY,
            Name TEXT
          )")

dbExecute(db.conn,"
          CREATE TABLE causes  
          (
            Cid INTEGER PRIMARY KEY,
            Name TEXT,
            Metric TEXT
          )")

dbExecute(db.conn,"
          CREATE TABLE series  
          (
            Sid INTEGER PRIMARY KEY,
            Name TEXT,
            Metric TEXT
          )")

dbExecute(db.conn,"
          CREATE TABLE yearlyGDB  
          (
            GBDid INTEGER PRIMARY KEY,
            Cause INTEGER,
            Data REAL,
            Location INTEGER,
            Year Integer,
            FOREIGN KEY (Cause) REFERENCES causes(Cid),
            FOREIGN KEY (Location) REFERENCES locations(Lid),
            FOREIGN KEY (Year) REFERENCES years(Yid)
          )")

dbExecute(db.conn,"
          CREATE TABLE yearlyWB  
          (
            WBid INTEGER PRIMARY KEY,
            Series INTEGER,
            Data REAL,
            Location INTEGER,
            Year Integer,
            FOREIGN KEY (Series) REFERENCES series(Sid),
            FOREIGN KEY (Location) REFERENCES locations(Lid),
            FOREIGN KEY (Year) REFERENCES years(Yid)
          )")

dbWriteTable(db.conn, "locations", df.locations, overwrite = T)
dbWriteTable(db.conn, "causes", df.causes, overwrite = T)
dbWriteTable(db.conn, "series", df.series, overwrite = T)
dbWriteTable(db.conn, "years", df.years, overwrite = T)
dbWriteTable(db.conn, "yearlyWB", df.yearlyWB, overwrite = T)
dbWriteTable(db.conn, "yearlyGDB", df.yearlyGBD, overwrite = T)

```

Don't forget to disconnect!
```{r}
dbDisconnect(db.conn)
```

