---
title: "Model_DS_5110_Team_11_Project_SP_2023"
author: "Taylor (Tay) Goodwin"
date: "2023-04-20"
output: pdf_document
---

# References:

Healthcare spending and health outcomes: evidence from selected East African countries. Murad A Bein,1 Dogan Unlucan,2 Gbolahan Olowu,1 and Wagdi Kalifa2. African Health Sciences. 2017 Mar; 17(1): 247–254. https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5636241/

Mortality and life expectancy forecast for (comparatively) high mortality countries. Ahbab Mohammad Fazle Rabbi & Stefano Mazzuco. Genus volume 74, Article number: 18 (2018)..  https://genus.springeropen.com/articles/10.1186/s41118-018-0042-x

Diabetes and Chronic Kidney Disease. Centers for Disease Control and Prevention. U.S. Department of Health & Human Services. https://www.cdc.gov/diabetes/managing/diabetes-kidney-disease.html#:~:text=Both%20type%201%20and%20type%202%20diabetes%20can%20cause%20kidney%20disease.&text=Kidney%20diseases%20are%20the%209th,begin%20treatment%20for%20kidney%20failure

What Are the Socio-Economic Predictors of Mortality in a Society?. Wahab Adewuyi Adejumo1, Adetunji Raimi Tijani1, Sheriff Adesanyaonatola2, 1 Department of Insurance, the Oke-Ogun Polytechnic Saki, Oyo State, Nigeria.
2Department of Insurance, the Polytechnic Ibadan, Oyo State, Nigeria. Journal of Financial Risk Management > Vol.8 No.4, December 2019. https://www.scirp.org/journal/paperinformation.aspx?paperid=96881 


# Setup Libraries 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Metrics)
library(modelr)
library("RSQLite")
library(tidyr)
library(viridis)
library(maps)
library(ModelMetrics)
library(tidyr)
library(dplyr)
library(panelr)
library(plm)
library(readxl)
library(ggplot2)
library(tidyverse)
library(readr)
library(ggpubr)
library(naniar)
library(reshape2)
library(DBI)
library(RSQLite)
library(reshape2)
library(extrafont)
library(stringr)
library(showtext)
```

Run this chunk to connect to database & build all data frames:
```{r}
library("RSQLite")
library(tidyr)
library(dplyr)
library(maps)
library(ggplot2)
library(showtext)
library(stringr)
library(tidyverse)

db.conn <- dbConnect(RSQLite::SQLite(), dbname="worldHealth_v3.db")

# Full list of countries
health.full<- dbGetQuery(db.conn,   
"SELECT c.Cause, ye.Year, l.Location, y.Data, l.Region, l.G20, l.IncomeLevel FROM yearlyGBD AS y
JOIN locations AS l ON y.Location = l.lid
JOIN causes AS c ON y.Cause = c.cid
JOIN years AS ye ON y.Year = ye.yid")

#Get econ data for countries w/ 10% or less data missing
econ.10p<- dbGetQuery(db.conn,   
"SELECT ye.year, y.Data, l.location, s.Series , l.Region, l.G20, l.IncomeLevel FROM yearlyWB AS y
JOIN locations AS l ON y.Location = l.lid
JOIN years AS ye ON y.Year = ye.yid
JOIN series as s ON y.Series = s.sid")

# Get health data for countries w/ 10% or less data missing
health.10p <- dbGetQuery(db.conn,   
"SELECT c.Cause, ye.Year, l.Location, y.Data, l.Region, l.G20, l.IncomeLevel FROM yearlyGBD AS y
JOIN locations AS l ON y.Location = l.lid
JOIN causes AS c ON y.Cause = c.cid
JOIN years AS ye ON y.Year = ye.yid")
health.10p <- health.10p[health.10p$Location %in% econ.10p$Location,]

# Pivot econ data
econ.10p_pivot <- econ.10p %>% 
  group_by(Year, Location, Series) %>% 
  slice(1) %>% 
  pivot_wider(names_from = Series, values_from = Data)

```

# Tidying and Separating Training Datasets
```{r data rename and separate training datasets}

health_data <- health.10p
economic <- econ.10p_pivot
var_names <- c(colnames(economic))

health_2019 <- health_data %>%
  select(Cause, Data, Location, Year, Region) %>%
  filter(Cause == "Diabetes and kidney diseases, Deaths", Year == "2019", Region != "Other" ) %>%
    arrange(Location) %>%
  arrange(Year)

health_2019_Inc <- health_data %>%
  select(Cause, Data, Location, Year, Region) %>%
  filter(Cause == "Diabetes and kidney diseases, Incidence", Year == "2019", Region != "Other" ) %>%
    arrange(Location) %>%
  arrange(Year)

health_2019_Inc$Diabetes_Kidn_Incid <- health_2019_Inc$Data

```

# Subsetting and Organizing Data, Final Combine
```{r subsetting and arranging source datasets}

#deaths training data 
diabetes_kidn_deaths <- health_data %>%
  select(Location, Year, Data, Cause) %>%
  filter(Cause == "Diabetes and kidney diseases, Deaths", Year != 2019) %>%
  arrange(Location) %>%
  arrange(Year)

diabetes_kidn_deaths$Diabetes_Kidn_Deaths <- diabetes_kidn_deaths$Data

#prevalence
diabetes_kidn_prev <- health_data %>%
  select(Location, Year, Data, Cause) %>%
  filter(Cause == "Diabetes and kidney diseases, Prevalence", Year != 2019) %>%
  arrange(Location) %>%
  arrange(Year)

diabetes_kidn_prev$Diabetes_Kidn_Prev <- diabetes_kidn_prev$Data

#incidence
diabetes_kidn_incid <- health_data %>%
  select(Location, Year, Data, Cause) %>%
  filter(Cause == "Diabetes and kidney diseases, Incidence", Year != 2019) %>%
  arrange(Location) %>%
  arrange(Year)

diabetes_kidn_incid$Diabetes_Kidn_Incid <- diabetes_kidn_incid$Data



#deaths testing data 
diabetes_kidn_deaths_2019 <- health_data %>%
  select(Location, Year, Data, Cause) %>%
  filter(Cause == "Diabetes and kidney diseases, Deaths", Year == 2019) %>%
  arrange(Location) %>%
  arrange(Year)

diabetes_kidn_deaths_2019$Diabetes_Kidn_Deaths <- diabetes_kidn_deaths_2019$Data

#prevalence
diabetes_kidn_prev_2019<- health_data %>%
  select(Location, Year, Data, Cause) %>%
  filter(Cause == "Diabetes and kidney diseases, Prevalence", Year == 2019) %>%
  arrange(Location) %>%
  arrange(Year)

diabetes_kidn_prev_2019$Diabetes_Kidn_Prev <- diabetes_kidn_prev_2019$Data


#incidence
diabetes_kidn_incid_2019 <- health_data %>%
  select(Location, Year, Data, Cause) %>%
  filter(Cause == "Diabetes and kidney diseases, Incidence", Year == 2019) %>%
  arrange(Location) %>%
  arrange(Year)

diabetes_kidn_incid_2019$Diabetes_Kidn_Incid <- diabetes_kidn_incid_2019$Data


econ_2019 <- economic %>%
  filter(Year == "2019")

econ_train_data <- economic %>%
  filter(Year != 2019)

deaths_2019 <-cbind(diabetes_kidn_deaths_2019[, c(1:2,5)], Diabetes_Kidn_Prev = diabetes_kidn_prev_2019$Diabetes_Kidn_Prev,Diabetes_Kidn_Incid = diabetes_kidn_incid_2019$Diabetes_Kidn_Incid)

deaths <-cbind(diabetes_kidn_deaths[, c(1:2,5)], Diabetes_Kidn_Prev = diabetes_kidn_prev$Diabetes_Kidn_Prev,Diabetes_Kidn_Incid = diabetes_kidn_incid$Diabetes_Kidn_Incid)

deaths_test <- deaths_2019 %>% 
  filter(Year == 2019)

df_2019 <- cbind(econ_2019, deaths_test)
df_2019 <- df_2019 %>% filter (Region != "Other")

df <- cbind(deaths[,c(3:5)], econ_train_data)
df <- df[df$Region != "Other",]
df <- df[df$Year != "2019",]

```

# Scatter plots for Single Linear Regression Models
```{r exploratory analysis - diabetes death rates for all years in testing data}

#  Death Rates and Incidence Rates
ggplot(data = df, aes(x = Diabetes_Kidn_Incid, color = as.factor(IncomeLevel), y = Diabetes_Kidn_Deaths)) + 
  geom_point() + 
  layer(geom="line",
        stat="smooth",
        position="identity", 
        params=list(method="auto", se=TRUE)) +
  labs(title = "Diabetes and Kidney Disease Incidence and Death Rates",
       subtitle = "Proportion of People Developed Diabetes/ Kidney Disease",
       x = "Diabetes and Kidney Disease Incidence Rates", y = "Diabetes and Kidney Disease Death Rates") + 
  facet_wrap(~ Year) +
  theme_minimal()


# Death Rates and Internet Access
# Inte
ggplot(data = df, aes(x = `Individuals using the Internet (% of population)`, color = as.factor(IncomeLevel), y = Diabetes_Kidn_Deaths)) + 
  geom_point() + 
    layer(geom="line",
        stat="smooth",
        position="identity",
        params=list(method="auto", se=TRUE)) +
  labs(title = "Internet Access as a Proxy for National Technological Access", 
       subtitle = "Measured as Individuals using the Internet (% of population)",
       x = "Proportion of People with Internet Access", y = "Diabetes and Kidney Disease Death Rates") +
  facet_wrap(~ Year) +
  theme_minimal()


# Death Rates and Gov't Health Expenditure
# In examining a scatter plot for domestic gov't health exp against diabetes caused deaths, skewness is apparent towards right side of the graph, so a log transformation is introduced.
ggplot(data = df, aes(x = `Domestic general government health expenditure per capita (current US$)`, color = as.factor(IncomeLevel), y = Diabetes_Kidn_Deaths)) + 
    geom_point() + 
  labs(title = "Gov't Health Expenditure and Diabetes & Kidney Disease Deaths", 
       subtitle= "General Government Health Expenditure per capita in US 2015 Dollars",
       x = "Domestic general government health expenditure per capita (current US$)", y= "Diabetes and Kidney Disease Death Rates") +
  layer(geom="line",
        stat="smooth",
        position="identity",
        params=list(method="auto", se=TRUE)) +
  facet_wrap(~ Year)

# Death Rates and Gov't Health Expenditure Continued 
# Log(govt health exp) 
ggplot(data = df, aes(x = log(`Domestic general government health expenditure per capita (current US$)`), color = as.factor(IncomeLevel), y = Diabetes_Kidn_Deaths)) + 
  labs(title = "Gov't Health Expenditure per capita and Diabetes Death Rates (log)", 
       subtitle = "add here",
       x = "Domestic general government health expenditure per capita (current US$)", y = "Diabetes and Kidney Disease Death Rates") +
  geom_point() + 
  layer(geom="line",
        stat="smooth",
        position="identity",
        params=list(method="auto", se=TRUE)) +
  facet_wrap(~ Year)

```

# Single linear regression models and correlation  - training data
```{r single linear models and correlation - training data}

df_2015<- df %>%
  filter(Year == 2015)


print("Correlation Coefficients:")
print(cor(df_2015$Diabetes_Kidn_Incid, df_2015$Diabetes_Kidn_Deaths))
print(cor(df_2015$`People using at least basic drinking water services (% of population)`, df_2015$Diabetes_Kidn_Deaths))
print(cor(df_2015$`People using at least basic sanitation services (% of population)`, df_2015$Diabetes_Kidn_Deaths))
print(cor(df_2015$`Total greenhouse gas emissions (kt of CO2 equivalent)`, df_2015$Diabetes_Kidn_Deaths))
print(cor(df_2015$`CO2 emissions (metric tons per capita)`, df_2015$Diabetes_Kidn_Deaths))
print(cor(df_2015$`Individuals using the Internet (% of population)`, df_2015$Diabetes_Kidn_Deaths))
print(cor(df_2015$`Proportion of seats held by women in national parliaments (%)`, df_2015$Diabetes_Kidn_Deaths))
print(cor(df_2015$`Domestic general government health expenditure per capita (current US$)`, df_2015$Diabetes_Kidn_Deaths))
print(cor(df_2015$`Domestic private health expenditure per capita (current US$)`, df_2015$Diabetes_Kidn_Deaths))

print("RMSE for Single Linear Regressions:")
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid))
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$IncomeLevel))
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`CO2 emissions (metric tons per capita)`))
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`Total greenhouse gas emissions (kt of CO2 equivalent)`))
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`Individuals using the Internet (% of population)`))
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`People using at least basic drinking water services (% of population)`))
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`Access to electricity (% of population)`))
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`Proportion of seats held by women in national parliaments (%)`))
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`Domestic general government health expenditure per capita (current US$)`))
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`Domestic private health expenditure per capita (current US$)`))

summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$IncomeLevel))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`Individuals using the Internet (% of population)`))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`CO2 emissions (metric tons per capita)`))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`Total greenhouse gas emissions (kt of CO2 equivalent)`))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`People using at least basic drinking water services (% of population)`))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`Access to electricity (% of population)`))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`Proportion of seats held by women in national parliaments (%)`))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`Domestic general government health expenditure per capita (current US$)`))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$`Domestic private health expenditure per capita (current US$)`))

```


# Multivariable linear regression model - training data 
```{r multivar regression - training data }

df_2015$log_Govt_Exp <- log(df_2015$`Domestic general government health expenditure per capita (current US$)`)
df_2019$log_Govt_Exp <- log(df_2019$`Domestic general government health expenditure per capita (current US$)`)

print("RMSE for Each Multivariable Model (Stepwise Composition):")

# control for the rate of development of diabetes/kidney disease cases in a year. Some cases may be missed and not caught in this variable and remain in  the noise, due to deaths by diabetes or kidney disease that go undetected. 
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid))
summary((lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid)))

# looking to asses any potential relationship between a nation's wealth level and deaths from diabetes and kidney disease. In addition do countries of different income levels have differing relationships to the dependent variable. 
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid + df_2015$IncomeLevel))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid + df_2015$IncomeLevel))

# aiming to control for any impact that a nations level of tehcnological advancement may have on death rates from diabetes and kidney disease. 
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid + df_2015$IncomeLevel + df_2015$`Individuals using the Internet (% of population)`))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid + df_2015$IncomeLevel + df_2015$`Individuals using the Internet (% of population)`))

# hoping to discern if environmental differences, such as pollution have a discernable impact on diabetes and kidney disease deaths. 
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid + df_2015$IncomeLevel + df_2015$`Individuals using the Internet (% of population)` + df_2015$`CO2 emissions (metric tons per capita)`))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid + df_2015$IncomeLevel + df_2015$`CO2 emissions (metric tons per capita)`))

# assessing whether governmental
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid + df_2015$IncomeLevel + df_2015$`Individuals using the Internet (% of population)` + df_2015$log_Govt_Exp))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid + df_2015$IncomeLevel + df_2015$log_Govt_Exp))



# Final Model Options for Consideration
# Model 1
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid + df_2015$IncomeLevel + df_2015$`Individuals using the Internet (% of population)`))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid + df_2015$IncomeLevel + df_2015$`Individuals using the Internet (% of population)`))
AIC(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid + df_2015$IncomeLevel + df_2015$`Individuals using the Internet (% of population)`))
plot(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid + df_2015$IncomeLevel + df_2015$`Individuals using the Internet (% of population)`))

# Model 2
ModelMetrics::rmse(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid  + df_2015$IncomeLevel + df_2015$`Individuals using the Internet (% of population)` + df_2015$log_Govt_Exp))
summary(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid +  df_2015$IncomeLevel +  df_2015$`Individuals using the Internet (% of population)` + df_2015$log_Govt_Exp))
AIC(lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid + df_2015$IncomeLevel+ df_2015$`Individuals using the Internet (% of population)`+ df_2015$log_Govt_Exp))
plot (lm(df_2015$Diabetes_Kidn_Deaths ~ df_2015$Diabetes_Kidn_Incid + df_2015$IncomeLevel+ df_2015$`Individuals using the Internet (% of population)`+ df_2015$log_Govt_Exp))

```

# Testing the 2015 trained model with 2019 predictions
```{r muti variable testing and predictive ability of model}

# Comparing Models Using 2019 Values and performing model validation
# Model 1
ModelMetrics::rmse(lm(df_2019$Diabetes_Kidn_Deaths ~ df_2019$Diabetes_Kidn_Incid + df_2019$IncomeLevel + df_2019$`Individuals using the Internet (% of population)`))
summary(lm(df_2019$Diabetes_Kidn_Deaths ~ df_2019$Diabetes_Kidn_Incid + df_2019$IncomeLevel + df_2019$`Individuals using the Internet (% of population)`))
AIC(lm(df_2019$Diabetes_Kidn_Deaths ~ df_2019$Diabetes_Kidn_Incid + df_2019$IncomeLevel + df_2019$`Individuals using the Internet (% of population)`))
plot(lm(df_2019$Diabetes_Kidn_Deaths ~ df_2019$Diabetes_Kidn_Incid + df_2019$IncomeLevel + df_2019$`Individuals using the Internet (% of population)`))

# Model 2
ModelMetrics::rmse(lm(df_2019$Diabetes_Kidn_Deaths ~ df_2019$Diabetes_Kidn_Incid  + df_2019$IncomeLevel + df_2019$`Individuals using the Internet (% of population)` + df_2019$log_Govt_Exp))
summary(lm(df_2019$Diabetes_Kidn_Deaths ~ df_2019$Diabetes_Kidn_Incid +  df_2019$IncomeLevel +  df_2019$`Individuals using the Internet (% of population)` + df_2019$log_Govt_Exp))
AIC(lm(df_2019$Diabetes_Kidn_Deaths ~ df_2019$Diabetes_Kidn_Incid + df_2019$IncomeLevel+ df_2019$`Individuals using the Internet (% of population)`+ df_2019$log_Govt_Exp))
plot (lm(df_2019$Diabetes_Kidn_Deaths ~ df_2019$Diabetes_Kidn_Incid + df_2019$IncomeLevel+ df_2019$`Individuals using the Internet (% of population)`+ df_2019$log_Govt_Exp))


reg_2015 <-(lm(df_2019$Diabetes_Kidn_Deaths ~ df_2019$Diabetes_Kidn_Incid + df_2019$IncomeLevel + df_2019$`Individuals using the Internet (% of population)`))

predictions <- predict(reg_2015, df_2019)

error_info = data.frame(
  actual = df_2019$Diabetes_Kidn_Deaths,
  prediction = predictions
)

RMSE <- Metrics::rmse(predictions, df_2019$Diabetes_Kidn_Deaths)
cat("RMSE:", RMSE, "\n")

```
