---
title: "Build World Bank/Global Burden Disease Database"
---


```{r}
if("DBI" %in% rownames(installed.packages()) == FALSE) {
  install.packages("DBI")  
}
library("DBI")

if("RSQLite" %in% rownames(installed.packages()) == FALSE) {
  install.packages("RSQLite")  
}
library("sqldf")

if("sqldf" %in% rownames(installed.packages()) == FALSE) {
  install.packages("sqldf")  
}
library("sqldf")

if("readxl" %in% rownames(installed.packages()) == FALSE) {
  install.packages("readxl")  
}
library("readxl")
library("stringr")
library(tidyverse)
options(sqldf.driver = 'SQLite')
```


```{r}
health <- na.omit(read.csv("HealthData.csv"))
incident.prevalence <- na.omit(read.csv("IncidencePrevalence.csv"))
locationCodes <- data_frame(na.omit(read.csv("All Locations Hierarchies-Table 1.csv")))


incident.prevalence <- incident.prevalence %>% select(
  measure = measure_name,
  location = location_name,
  sex = sex_name,
  cause = cause_name,
  metric = metric_name,
  year,
  val
)

health <- health %>% select(
  measure,
  location,
  sex,
  cause,
  metric,
  year,
  val
)

locationCodes <- locationCodes %>% select(
  Location.ID,
  Location.Name, 
  Parent.ID,
  Level
)

health <- health[health$cause == "Cardiovascular diseases" | 
                   health$cause == "Diabetes and kidney diseases",]

healthFull <- rbind(health, incident.prevalence)
healthFull <- healthFull %>% mutate(cause = paste(paste0(cause, ","), measure))
healthFull$location <- str_squish(gsub("\\(.*", "", healthFull$location))
locationCodes$Location.ID <- str_squish(gsub("\\(.*", "", locationCodes$Location.ID))



```


Normalize country names in health data
```{r}
healthFull$location[healthFull$location == "Republic of Moldova"] <- "Moldova"
healthFull$location[healthFull$location == "United States of America"] <- "United States"
healthFull$location[healthFull$location == "Côte d'Ivoire"] <- "Cote d'Ivoire"
healthFull$location[healthFull$location =="Lao People's Democratic Republic"] <- "Lao PDR"
healthFull$location[healthFull$location =="Saint Kitts and Nevis"] <- "St. Kitts and Nevis"
healthFull$location[healthFull$location =="Saint Lucia"] <- "St. Lucia"
healthFull$location[healthFull$location =="Saint Vincent and the Grenadines"] <- "St. Vincent and the Grenadines"
healthFull$location[healthFull$location =="United Republic of Tanzania"] <- "Tanzania"
healthFull$location[healthFull$location =="Viet Nam"] <- "Vietnam"

locationCodes$Location.Name[locationCodes$Location.Name == "Republic of Moldova"] <- "Moldova"
locationCodes$Location.Name[locationCodes$Location.Name == "United States of America"] <- "United States"
locationCodes$Location.Name[locationCodes$Location.Name == "Côte d'Ivoire"] <- "Cote d'Ivoire"
locationCodes$Location.Name[locationCodes$Location.Name =="Lao People's Democratic Republic"] <- "Lao PDR"
locationCodes$Location.Name[locationCodes$Location.Name =="Saint Kitts and Nevis"] <- "St. Kitts and Nevis"
locationCodes$Location.Name[locationCodes$Location.Name =="Saint Lucia"] <- "St. Lucia"
locationCodes$Location.Name[locationCodes$Location.Name =="Saint Vincent and the Grenadines"] <- "St. Vincent and the Grenadines"
locationCodes$Location.Name[locationCodes$Location.Name =="United Republic of Tanzania"] <- "Tanzania"
locationCodes$Location.Name[locationCodes$Location.Name =="Viet Nam"] <- "Vietnam"

```

```{r}
WorldBankIncome <- locationCodes[locationCodes$Location.ID %in% c(44575:44578),]
G20 <- locationCodes[locationCodes$Parent.ID == 44586 & locationCodes$Location.ID != 44586,]
WorldBankRegion <- locationCodes[locationCodes$Parent.ID == 478 & locationCodes$Location.ID != 478,]

WorldBankIncomeCountries <- locationCodes[locationCodes$Parent.ID %in% WorldBankIncome$Location.ID,]
WorldBankRegionCountries <- locationCodes[locationCodes$Parent.ID %in% WorldBankRegion$Location.ID,]

WorldBankIncomeCountries
WorldBankRegionCountries
G20
```


Build data frames for each table
```{r}
db.conn <- dbConnect(RSQLite::SQLite(), dbname="worldHealth_v3.db")

df.yearlyGBD <- sqldf::sqldf(
  "select 1 as GBid,
  cause as Cause,
  val as Data,
  location as Location,
  year as Year
  from `healthFull`"
)

df.locations <- sqldf::sqldf("
  select 1 as lid, 
  Location,
  1 as IncomeLevel,
  1 as G20,
  1 as Region
  from `df.yearlyGBD` 
  group by Location
  ")

df.years <- sqldf::sqldf(
  "select 1 as yid,
  year as Year
  from `df.yearlyGBD`
  group by Year
  ")

df.causes <- sqldf::sqldf(
  "select 1 as cid,
  cause as Cause,
  metric as Metric
  from `healthFull`
  group by Cause"
)

```


Set foreign keys
```{r}
n.locations <- nrow(df.locations)
df.locations[,1] <- seq(1, n.locations)

n.years <- nrow(df.years)
df.years[,1] <- seq(1, n.years)

n.causes <- nrow(df.causes)
df.causes[,1] <- seq(1, n.causes)


n.yearlyGBD <- nrow(df.yearlyGBD)
df.yearlyGBD[,1] <- seq(1, n.yearlyGBD)


for(r in 1:n.yearlyGBD){
  # set location id
  lid <- df.locations$lid[which(df.locations$Location == df.yearlyGBD$Location[r])]
  df.yearlyGBD$Location[r] <- lid

  # set series id
  cid <- df.causes$cid[which(df.causes$Cause == df.yearlyGBD$Cause[r])]
  df.yearlyGBD$Cause[r] <- cid
  
  # set year id
  yid <- df.years$yid[which(df.years$Year == df.yearlyGBD$Year[r])]
  df.yearlyGBD$Year[r] <- yid

}

```



```{r}
for(country in df.locations$Location){
  row <- df.locations[df.locations$Location == country,]
  if(country %in% G20$Location.Name){
    row$G20 <- "TRUE"
  } else {
    row$G20 <- "FALSE"
  }
  if(country %in% WorldBankIncomeCountries$Location.Name){
    wbRow <- WorldBankIncomeCountries[WorldBankIncomeCountries$Location.Name == country,]
    row$IncomeLevel <-str_sub(WorldBankIncome$Location.Name[WorldBankIncome$Location.ID == 
                                                              wbRow$Parent.ID], start = 11)
  }
  if(country %in% WorldBankRegionCountries$Location.Name){
    wrRow <- WorldBankRegionCountries[WorldBankRegionCountries$Location.Name == country,]
    row$Region <- str_sub(WorldBankRegion$Location.Name[WorldBankRegion$Location.ID == 
                                                          wrRow$Parent.ID], end = -6)
  } else {
    row$Region <- "Other"
  }
  df.locations[df.locations$Location == country,] <- row;
}
df.locations
```


Write tables to db
```{r}
dbExecute(db.conn,"
          DROP TABLE IF EXISTS yearlyGBD")

dbExecute(db.conn,"
          DROP TABLE IF EXISTS years")
dbExecute(db.conn,"
          DROP TABLE IF EXISTS causes")
dbExecute(db.conn,"
          DROP TABLE IF EXISTS locations")

```

Write health data to DB
```{r}
dbExecute(db.conn,"
          CREATE TABLE years  
          (
            Yid INTEGER PRIMARY KEY,
            Name TEXT
          )")

dbExecute(db.conn,"
          CREATE TABLE locations  
          (
            Lid INTEGER PRIMARY KEY,
            IncomeLevel TEXT,
            G20 TEXT,
            Region,
            Name TEXT
          )")

dbExecute(db.conn,"
          CREATE TABLE causes  
          (
            Cid INTEGER PRIMARY KEY,
            Name TEXT,
            Metric TEXT
          )")

dbExecute(db.conn,"
          CREATE TABLE yearlyGBD 
          (
            GBDid INTEGER PRIMARY KEY,
            Cause INTEGER,
            Data REAL,
            Location INTEGER,
            Year Integer,
            FOREIGN KEY (Cause) REFERENCES causes(Cid),
            FOREIGN KEY (Location) REFERENCES locations(Lid),
            FOREIGN KEY (Year) REFERENCES years(Yid)
          )")


dbWriteTable(db.conn, "locations", df.locations, overwrite = T)
dbWriteTable(db.conn, "causes", df.causes, overwrite = T)
dbWriteTable(db.conn, "years", df.years, overwrite = T)
dbWriteTable(db.conn, "yearlyGBD", df.yearlyGBD, overwrite = T)

```
# Econ data starts here
```{r} 
WB_DBI <- na.omit(read.csv("WB_WDI.csv"))
worldBank <- WB_DBI %>% select(
                  location = Country.Name, 
                  series = Series.Name,
                  `2000` = X2000..YR2000., 
                  `2005` = X2005..YR2005.,
                  `2010` = X2010..YR2010.,
                  `2015` = X2015..YR2015.,
                  `2019` = X2019..YR2019.) %>%
  pivot_longer(`2000`:`2019`, names_to = "year") 

worldBank$value[worldBank$value == ".."] <- NA

worldBank$location <- str_squish(gsub("\\(.*", "", worldBank$location))
worldBank$value <- as.numeric(worldBank$value)
worldBank
```

Normalize names in economic data:
```{r}
worldBank$location[worldBank$location == "Bahamas, The"] <- "Bahamas"
worldBank$location[worldBank$location == "Korea, Dem. People's Rep."] <- "Democratic People's Republic of Korea"
worldBank$location[worldBank$location == "Korea, Rep."] <- "Republic of Korea"
worldBank$location[worldBank$location == "Congo, Dem. Rep."] <- "Democratic Republic of the Congo"
worldBank$location[worldBank$location == "Egypt, Arab Rep."] <- "Egypt" 
worldBank$location[worldBank$location == "Gambia, The"] <- "Gambia"
worldBank$location[worldBank$location == "Iran, Islamic Rep."] <- "Iran"
worldBank$location[worldBank$location == "Kyrgyz Republic"] <- "Kyrgyzstan"
worldBank$location[worldBank$location == "Congo, Rep."] <- "Congo"
worldBank$location[worldBank$location == "Micronesia, Fed. Sts."] <- "Micronesia"
worldBank$location[worldBank$location == "West Bank and Gaza"] <- "Palestine"
worldBank$location[worldBank$location == "Slovak Republic"] <- "Slovakia"
worldBank$location[worldBank$location == "Turkiye"] <- "Turkey"
worldBank$location[worldBank$location == "Virgin Islands"] <- "United States Virgin Islands"
worldBank$location[worldBank$location == "Venezuela, RB"] <- "Venezuela"
worldBank$location[worldBank$location == "Yemen, Rep."] <- "Yemen"
```

These names are in econ data but not in health data
```{r}
countriesW <- unique(worldBank$location)
countriesH <- unique(healthFull$location)
wcountries <- countriesW[!countriesW %in% countriesH]
wcountries[order(wcountries)]
```

These names are in health but not in econ data
```{r}
countriesW <- unique(worldBank$location)
countriesH <- unique(healthFull$location)
hcountries <- countriesH[!countriesH %in% countriesW]
hcountries[order(hcountries)]

```

Get rid of countries not in the health data
```{r}
countriesW <- unique(worldBank$location)
countriesH <- unique(healthFull$location)
wcountries <- countriesW[!countriesW %in% countriesH]

worldBank <- worldBank[!(worldBank$location %in% wcountries),]
```

Count how many nulls each Series has and the percent missing of total series
```{r}
series <- unique(worldBank$series[worldBank$series != ""])

null.count <- data.frame(Series = character(length(series)))
for(j in 1:length(series)){
  null.count$NullCount[j] <- 0
  null.count$PercentMissing[j] <- 0
}

j = 1
for(i in 1:length(series)){
  col <-worldBank[worldBank$series == series[i],]
  null.count$Series[i] = series[i]
  null.count$NullCount[i] = sum(is.na(col$value))
  null.count$PercentMissing[i] = sum(is.na(col$value)) / nrow(col) * 100

}

null.count <- null.count[order(null.count$NullCount),]
null.count
```


Eliminate all series with more than 10% missing:
```{r}
seriesList <- null.count$Series[null.count$PercentMissing > 10]
seriesList

economic.10p <- worldBank[!worldBank$series %in% seriesList,]
economic.10p
```

Count how many nulls each country has and the percent missing
```{r}
locations <- unique(economic.10p$location[economic.10p$location != ""])

null.count1 <- data.frame(Countries = character(length(locations)))
for(j in 1:length(locations)){
  null.count1$NullCount[j] <- 0
  null.count1$PercentMissing[j] <- 0
}


for(i in 1:length(locations)){
  col <-economic.10p[economic.10p$location == locations[i],]
  null.count1$Countries[i] = locations[i]
  null.count1$NullCount[i] = sum(is.na(col$value))
  null.count1$PercentMissing[i] = sum(is.na(col$value)) / nrow(col) * 100
}

null.count1 <- null.count1[order(null.count1$NullCount),]
null.count1
```

Total missing:
With these 200 countries, 4.25% of our data values are null, but there are a few countries that are missing A LOT of data points.
If countries are each an observation, I imagine we could eliminate some as long as it doesn't skew the data in some specific direction
```{r}
sum(is.na(economic.10p$value)) / length(economic.10p$value) * 100
```


Eliminate all countries with more than 10% missing:
```{r}
locList <- null.count1$Countries[null.count1$PercentMissing > 10]
locList

economic.10p <- economic.10p[!economic.10p$location %in% locList,]
health <- health[!health$location %in% locList,]
```

Total percentage of nulls goes down to less than 1%
```{r}
sum(is.na(economic.10p$value)) / length(economic.10p$value) * 100
```

# Impute NA values:
```{r}
economic.10p[is.na(economic.10p$value),]
mean_values <- ave(as.numeric(economic.10p$value), economic.10p$location, economic.10p$series, FUN = function(x) mean(x, na.rm=TRUE))

economic_i <- economic.10p
economic_i$value <- ifelse(is.na(economic_i$value), mean_values, economic_i$value)

```


```{r}
# these rows were unable to be imputed as all years were null, we'll eliminate these countries from the data set as well
economic_i[economic_i$value == "NaN",]
countries_del <- unique(economic_i$location[economic_i$value == "NaN"])

economic_i <- economic_i[!economic_i$location %in% countries_del,]
countries_del
```



# Econ database build starts here
Build data frames for each table
```{r}
db.conn <- dbConnect(RSQLite::SQLite(), dbname="worldHealth_v3.db")

df.yearlyWB <- sqldf::sqldf("
  select 1 as WBid,
  series as Series,
  value as Data,
  location as Location,
  year as Year
  from `economic_i`")

df.series <- sqldf::sqldf(
  "select 1 as sid,
  series as Series,
  series as Metric
  from `economic_i`
  group by Series"
)

```


Set foreign keys
```{r}

n.series <- nrow(df.series)
df.series[,1] <- seq(1, n.series)

n.yearlyWB <- nrow(df.yearlyWB)
df.yearlyWB[,1] <- seq(1, n.yearlyWB)

for(j in 1:n.yearlyWB){
  # set location id
  lid <- df.locations$lid[which(df.locations$Location == df.yearlyWB$Location[j])]
  df.yearlyWB$Location[j] <- lid
    
  # set series id
  sid <- df.series$sid[which(df.series$Series == df.yearlyWB$Series[j])]
  df.yearlyWB$Series[j] <- sid
  
  # set year id
  yid <- df.years$yid[which(df.years$Year == df.yearlyWB$Year[j])]
  df.yearlyWB$Year[j] <- yid

}

```


```{r}
df.yearlyWB
```


Write tables to db
```{r}

dbExecute(db.conn,"
          DROP TABLE IF EXISTS yearlyWB")

dbExecute(db.conn,"
          DROP TABLE IF EXISTS series")



dbExecute(db.conn,"
          CREATE TABLE series  
          (
            Sid INTEGER PRIMARY KEY,
            Name TEXT,
            Metric TEXT
          )")

dbExecute(db.conn,"
          CREATE TABLE yearlyWB  
          (
            WBid INTEGER PRIMARY KEY,
            Series INTEGER,
            Data REAL,
            Location INTEGER,
            Year Integer,
            FOREIGN KEY (Series) REFERENCES series(Sid),
            FOREIGN KEY (Location) REFERENCES locations(Lid),
            FOREIGN KEY (Year) REFERENCES years(Yid)
          )")

dbWriteTable(db.conn, "series", df.series, overwrite = T)
dbWriteTable(db.conn, "yearlyWB", df.yearlyWB, overwrite = T)

```



Don't forget to disconnect!
```{r}
dbDisconnect(db.conn)
```

